---
title: "Lectures"
---


```{r include = FALSE}
library(dplyr)
library(stringr)
library(httr)
```



```{r include = FALSE}
# Import schedule
rawdat <- read.csv(file.path("course-admin", "schedule_2023.csv"), stringsAsFactors = FALSE, colClasses = list(course = 'factor', dow = 'factor'))
rawdat$course <- as.factor(rawdat$course)

# Class meetings
rawdat_cm <- rawdat[rawdat$course == "STAT540",]

cdat <- rawdat_cm %>%
  mutate(ugly_date = ISOdate(2022, match(month, month.name), day),
         date = sprintf("%s %s %d",
                        dow, months(ugly_date, abbreviate = TRUE),
                        day),
         notes = ifelse(grepl("^lect", lect) & raw_notes != "",
                        sprintf("%s: %s", lect, raw_notes), raw_notes))
cdat %>% head(3)


## link to the class page of any class meetings where such exists
req <- GET("https://api.github.com/repos/STAT540-UBC/lectures/git/trees/main?recursive=1")
stop_for_status(req)
filelist <- unlist(lapply(content(req)$tree, "[", "path"), use.names = F)

# now grab pdf file names
cmdat <-
  data.frame(pdf = filelist[grepl('^lect.*pdf', basename(filelist))],
             stringsAsFactors = FALSE)

cmdat <- cmdat %>% mutate(lect = str_extract(pdf, "lect[0-9]+"))
cmdat$lect <- sapply(cmdat$lect, str_replace, pattern="lect0*([0-9]+)", replacement="lecture-\\1")

if (nrow(cmdat) >0){
  cdat <- cdat %>% left_join(cmdat, by="lect")
}else{
  cdat$pdf <- NA
}

pdf_exists <- Negate(is.na)(cdat$pdf)

# Insert placeholder link to notes 
#cdat$notes[pdf_exists] <-
 # str_replace(cdat$notes[which(pdf_exists)],
#              "(lecture-[0-9]+)|(lecture-[1-9]+)", '<a href="LINK">\\1</a>')
cdat$notes <- gsub("lecture-", "Lecture ", cdat$notes)

cdat$notes[pdf_exists] <-
  str_replace(cdat$notes[which(pdf_exists)],
              "(Lecture [0-9]+)|(Lecture [1-9]+)", '[\\1](LINK)')
# Replace placeholder "LINK" which pdf file names
if (sum(pdf_exists) > 0)
  cdat$notes[pdf_exists] <- str_replace(cdat$notes[which(pdf_exists)], "LINK", paste0("https://github.com/STAT540-UBC/lectures/raw/main/", cdat$pdf[which(pdf_exists)]))


dat_cm <- cdat %>% split(cdat$course)
dat_cm <- lapply(dat_cm, 'rownames<-', NULL)
```



### Class meetings and schedule 

**Time** : Mon Wed 9:30 - 11am

**Location**: 

* *Before February 7*: as part of UBC's response to the COVID-19 omicron variant, class will be held online for the first part of the term (synchronously via Zoom; meeting links posted in Canvas)
* *On February 7 and after*: it is expected that class will be held in person in [ESB 2012](https://ssc.adm.ubc.ca/classroomservices/function/viewlocation?userEvent=ShowLocation&buildingID=ESB&roomID=2012) after the online transition period. In case of any changes to this plan, stay tuned to [UBC Broadcast communications](https://covid19.ubc.ca/) 


**Supplemental Files**: 

* Source files for lectures created using Rmarkdown can be found the [STAT540-UBC/lectures GitHub repo](https://github.com/STAT540-UBC/lectures)
* Additional companion notes can be found in the [STAT540-UBC/resources GitHub repo](https://github.com/STAT540-UBC/resources)

```{r results = 'asis', echo = FALSE}
knitr::kable(dat_cm[["STAT540"]] %>% select(date, notes, instructor) %>%
  dplyr::rename(Date=date, Topic=notes, Instructor=instructor),
  "html", escape = FALSE) %>% 
  kableExtra::column_spec(1, width = "6em") %>%
  kableExtra::kable_styling(full_width = TRUE)
```

