---
title: "Seminars"
---


```{r include = FALSE}
library(dplyr)
library(stringr)
library(glue)
library(here)
```

```{r include = FALSE}
githubLink = 'https://github.com/STAT540-UBC/STAT540-UBC.github.io'
seminarPathGithubIO = 'seminars'

# Import schedule
rawdat <- read.csv(file.path("course-admin", "schedule_2022.csv"), stringsAsFactors = FALSE, colClasses = list(course = 'factor', dow = 'factor'))
rawdat$course <- as.factor(rawdat$course)

# Seminar meetings 
rawdat_sm <- rawdat[rawdat$course == "STAT540-Seminar",]

sdat <- rawdat_sm %>%
  mutate(ugly_date = ISOdate(2021, match(month, month.name), day),
         date = sprintf("%s %s %d",
                        dow, 
                        months(ugly_date, abbreviate = TRUE),
                        day))
sdat %>% head(3)

## link to the page of any seminar where such exists
smdat <-
  data.frame(html = list.files(path=glue("../{seminarPathGithubIO}"), pattern = '^sm.*(\\.Rmd)$', 
                               full.names=TRUE,recursive = TRUE))
smdat <- smdat %>% mutate(lect = str_extract(html, "sm[0-9]+[a-z]*"))
smdat$lect <- sapply(smdat$lect, str_replace, pattern="sm0*([0-9]+)([a-z]*)", replacement="seminar-\\1\\2")

if(nrow(smdat) > 0){
  sdat <- sdat %>% left_join(smdat, by="lect")
}else{
  sdat$html <- NA
}

# repeat for md
smdat2 <-  data.frame(md = list.files(path=glue("../{seminarPathGithubIO}"), pattern = '^sm.*(\\.md)$', 
                               full.names=TRUE,recursive = TRUE))
smdat2 <- smdat2 %>% mutate(lect = str_extract(md, "sm[0-9]+[a-z]*"))
smdat2$lect <- sapply(smdat2$lect, str_replace, pattern="sm0*([0-9]+)([a-z]*)", replacement="seminar-\\1\\2")

if(nrow(smdat2) > 0){
  sdat <- sdat %>% left_join(smdat2, by="lect")
}else{
  sdat$md <- NA
}

html_exists = Negate(is.na)(sdat$html)
md_exists = Negate(is.na)(sdat$md)

sdat$lect[html_exists] <-
  str_replace(sdat$lect[html_exists],
              pattern="(^seminar-.*)", replacement='[\\1](LINK)')
sdat$lect[html_exists] <- str_replace(sdat$lect[html_exists], 
                                      "LINK", 
                                      file.path(githubLink,'blob/master',
                                                str_replace(sdat$html[html_exists],'\\.\\.','')
                                                ))

sdat$md[md_exists] <- file.path(githubLink,'blob/master',
          str_replace(sdat$md[md_exists],'\\.\\.',''))
                                                

#sdat$lect[html_exists] <-
#  str_replace(sdat$lect[html_exists],
#              pattern="(^seminar-.*)", replacement='<a href="LINK">\\1</a>')
#sdat$lect[html_exists] <- str_replace(sdat$lect[html_exists], 
#                                      "LINK", 
#                                      file.path(githubLink,'tree/master',
#                                                str_replace(sdat$html[html_exists],'\\.\\.','')
#                                                ))

#sdat$lect <- gsub("https://github.com/STAT540-UBC/STAT540-UBC.github.io/blob/master/", "..", sdat$lect)

dat_sm <- sdat %>% split(sdat$course)
dat_sm <- lapply(dat_sm, 'rownames<-', NULL)
```


### Seminars

**Time**: Monday 12 - 1pm

**Location**: 

* *Before January 24*: as part of UBC's response to the COVID-19 omicron variant, seminar will be held online for the first two weeks (synchronously via Zoom; meeting links posted in Canvas)
* *On January 24 and after*: it is expected that seminar will be held in person in [ESB 1012](https://ssc.adm.ubc.ca/classroomservices/function/viewlocation?userEvent=ShowLocation&buildingID=ESB&roomID=1012) after the first two weeks. In case of any changes to this plan, stay tuned to [UBC Broadcast communications](https://covid19.ubc.ca/) 


We strongly recommend reading the seminar materials prior to attending each seminar

```{r results = 'asis', echo = FALSE}
assignment_data <- read.csv("course-admin/assn_2022.csv",as.is = T,sep = ",")
seminar_num <- assignment_data$content
seminar_num[!grepl("Seminar", seminar_num)] <- NA
seminar_num <- gsub("Seminar ", "", seminar_num)
seminar_num <- gsub(" &", "", seminar_num)


htmls <- dat_sm[["STAT540-Seminar"]]$html

if (sum(is.na(htmls)) < length(htmls)){
 htmls <- tolower(sapply(strsplit(htmls, "/"), 
                          function(x) if(length(x) >     1){x[[4]]}else{x}))
 htmls <- gsub("seminar", "", htmls)
 htmls <- gsub("10", "X", htmls)
}

dat_sm[["STAT540-Seminar"]] <- dat_sm[["STAT540-Seminar"]] %>% 
  mutate(Rmd = lect) %>%
  mutate(both = ifelse(md_exists, paste0(Rmd, " ([md](", md, "))"), 
                       Rmd))


knitr::kable(dat_sm[["STAT540-Seminar"]] %>% select(date, both, raw_notes, instructor, mode) %>%
            dplyr::rename(Date=date, `Rmd (md)`= both,
                          Topic=raw_notes, 
                          Instructor=instructor, Mode=mode), 
            "html") %>% 
  kableExtra::column_spec(1, width = "6em") %>%
  kableExtra::kable_styling(full_width = TRUE)
```

**Supplemental materials** 

  * seminar 0a | Getting Started with R ([PDF](https://github.com/STAT540-UBC/STAT540-UBC.github.io/blob/master/seminars/seminar0a/getting_started_with_R.pdf), [Worksheet](https://github.com/STAT540-UBC/STAT540-UBC.github.io/raw/master/seminars/seminar0a/Seminar%200a%20-%20Getting%20Started.docx))
  * seminar 0b | Molecular Biology Intro ([PDF](https://github.com/STAT540-UBC/STAT540-UBC.github.io/raw/master/seminars/seminars_legacy/sm01e_biology-intro-2017.pdf), [Quiz to check your understanding in Canvas - not for a grade](https://canvas.ubc.ca/courses/61673/assignments/785665))
  