---
title: "Syllabus"
description: |
  STAT 540: Statistical Methods for High Dimensional Biology
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

Winter Term 2 2022 (January 10, 2022 - April 8, 2022) 

STAT 540 is a 3 credit course with a mandatory computing seminar 

Cross-listed as STAT 540, BIOF 540, GSAT 540

## Land acknowledgement

We respectfully recognize that the University of British Columbia Vancouver campus is located on the traditional, ancestral, and unceded territory of the xʷməθkʷəy̓əm (Musqueam) people. As we begin the term in a virtual learning environment, we are coming together from locations throughout Vancouver, BC, Canada, and the world. Please take a moment to learn about the territory you are occupying by visiting [this interactive indigenous land map](https://native-land.ca/).

## Course-level learning objectives

  * Perform exploratory data analysis and visualize genomics data
  * Apply tailored statistical methods to answer questions using high dimensional biological data
  * Make your work reproducible, reusable, and shareable
  * Work with real data in a collaborative model

## Selected topics 

  * Basics of molecular genetics/genomic and data collection assays (methods)
  * Basic probability and math foundations
  * Exploratory data analysis and data quality control
  * Normalization, batch correction
  * Causal inference and confounding effects
  * Basic statistical inference (“one gene at a time”) – linear models
  * Large-scale inference (“genome-wide”) – multiple testing
  * Analysis of microarray, RNASeq, and epigenetics data
  * Principal Component Analysis and clustering (unsupervised machine learning)
  * Classification and cross validation (supervised machine learning)
  * Gene set analysis and gene networks
  * Genome-wide association analysis (GWAS) 
  * Single-cell genomics

## Teaching Team

For more info on the Teaching team, including brief bios, see the 'People' pages on this website (linked below).

### Instructors

* [Keegan Korthauer, PhD](about_keegan.html) (She/Her/Hers)  
Email: <keegan@stat.ubc.ca>  
Virtual office hours: Thursday 9-10am (Pacific time)

* [Yongjin Park, PhD](about_yongjin.html) (He/Him/His)  
Email: <ypp@stat.ubc.ca>  
Virtual office hours: Wednesday 1-2pm (Pacific time)

### Teaching Assistants

* [Asfar Lathif](about_asfar.html) (He/Him/His)  
Email: <asfarlathifbt@gmail.com>  
Virtual office hours: Tuesday 10-11am (Pacific time)  

* [Marco Tello](about_marco.html) (He/Him/His)  
Email: <Marco.TelloPalencia@bcchr.ca>  
Virtual office hours: Thursday 11am-12pm TBA (Pacific time)

## Schedule  

Note that as part of UBC's response to the COVID-19 omicron variant, lecture and seminar will be held online for the first two weeks of the term (synchronously via Zoom; meeting links posted in Canvas). It is expected that lecture and seminar will be held in person starting January 24. In case of any changes to this plan, stay tuned to [UBC Broadcast communications](https://covid19.ubc.ca/).

Lectures held in person will also be live-streamed whenever possible (links shared in Canvas) to minimize disruption for those students who are unable to attend class due to illness or isolation. Seminar sessions will not be recorded, as these are primarily interactive.

### Lectures (Sec 201)

- Time : Mon Wed 9:30 - 11am
- Location: 
    * *Before January 24*: online (Zoom meeting links posted in Canvas)
    * *On January 24 and after*: in person in [ESB 2012](https://ssc.adm.ubc.ca/classroomservices/function/viewlocation?userEvent=ShowLocation&buildingID=ESB&roomID=2012) 
- See [Lectures](lectures.html) for lecture materials and schedule 

### Seminars (Sec S2B) 

- Time : Mon 12pm - 1pm
- Location: 
    * *Before January 24*: online (Zoom meeting links posted in Canvas)
    * *On January 24 and after*: in person in [ESB 1012](https://ssc.adm.ubc.ca/classroomservices/function/viewlocation?userEvent=ShowLocation&buildingID=ESB&roomID=1012)

- See [Seminars](seminars.html) for schedule and seminar materials 
- We strongly recommend reading the seminar materials prior to attending each seminar. 

## Course communication

As we begin the term with web-oriented learning, it is vital that we have a dedicated plan for how we will communicate throughout the semester.

#### Announcements 

Course announcements will be posted in the [course Github Discussion repository](https://github.com/STAT540-UBC-2022/Discussion) (this will be changed to a private repo and you will be granted access after we collect your Github user IDs). You are responsible for checking it regularly (subscribe to email notifications for this repo to make this easier).

#### General questions

Please also use the [course Github Discussion repository](https://github.com/STAT540-UBC-2022/Discussion) for posting questions (e.g. topics discussed in class, questions about course organization, assignment clarifications, if you are stuck on an assignment and need help). This ensures the message can be seen by the entire teaching team, and that others in the class who might have the same question can learn from responses. You are also welcome to share your input on questions posted by others using comments. 

#### Private matters

For private matters (e.g. requesting an extension or concerns about your grades), please contact the Teaching team by email (listed above). Please refrain from using email to ask general questions described above.

#### Group work

In your final project groups, we encourage you to make use of (1) the discussion feature in your GitHub Teams group, and (2) the issues feature in your group's project repo. In addition, you are encouraged to meet regularly via an online platform of your choice (or in person if conditions permit). Please reach out if you have any questions or challenges in this space.

## Prerequisites and Resources

This interdisciplinary course requires a broad range of skills at the interface of statistics, molecular biology / genomics, and computing. While you may have some background in one or more of the following areas, you are not expected to be an expert in all. That said, to be successful in the course, you may need to spend a bit more time in the areas that you have less experience in. Although there are no official prerequisites for the course, here is a list of useful skills to bring into the course and/or learn along the way.

### Statistics: 

- You should have already taken a university level introductory statistics course. 
- [This free online book "Modern Statistics for Modern Biology" by Susan Holmes and Wolfgang Huber](http://web.stanford.edu/class/bios221/book/) is a great reference for introduction or review of many of the statistical concepts that are relevant for this course.
- [This free online book "Data Analysis for the Life Sciences" by Rafael Irizarry and Michael Love](http://genomicsclass.github.io/book/) is another great resource for introduction or review of many of the statistical concepts relevant in this course, with an emphasis on implementation in R.

### Biology: 

- No requirements, but you are expected to learn things like, e.g. the difference between DNA and RNA, and the difference between a gene and a genome.
- See [this video](https://www.youtube.com/watch?v=lSqUDu4zb5k) and this [slideset](https://github.com/STAT540-UBC/resources/biology-intro-2017.pdf) for some basic introductory material. 
- Go through the (optional, not for a grade) molecular biology quiz on Canvas
- [This free online book "Concepts of Biology" by Fowler, Roush & Wise](https://openstax.org/books/concepts-biology/pages/1-introduction) is a great resource for biological concepts, in particular chapters 6 and 9
- [This free online book "Biology" by Clark, Douglas & Choi](https://openstax.org/books/biology-2e/pages/1-introduction) goes more in-depth, see Chapters 14, 15, and 16 for material on genetics that is particularly relevant for this course.
- No matter your prior experience, when you come across a new biological concept during the course or in your research, you might need to spend a bit of time 'learning as you go' - this is expected and something I still do regularly in my day-to-day research!

### R: 

- No experience required but be prepared to do a lot of self-guided learning if you haven't taken other courses on R or used it in your research.
- Start now by installing [R](https://cran.r-project.org/) and the HIGHLY RECOMMENDED "integrated development environment" (IDE) [RStudio](https://rstudio.com/products/rstudio/download/) - both are free and open source. 
- You should be able to run R on your own computer during each seminar session.
- If you are new to R, check out [this blog post on getting started with R](http://santina.me/Get-started-with-R/).
- [This free online book "Introduction to Data Science" by Rafael Irizarry](https://rafalab.github.io/dsbook/) is also a great resource for getting more in-depth with R, programming basics, and the tidyverse. In particular see Chapters 1-5:
  - [Chapter 1: Getting Started with R and R Studio](https://rafalab.github.io/dsbook/getting-started.html)
  - [Chapter 2: R Basics](https://rafalab.github.io/dsbook/r-basics.html)
  - [Chapter 3: Programming Basics](https://rafalab.github.io/dsbook/programming-basics.html)
  - [Chapter 4: The tidyverse](https://rafalab.github.io/dsbook/tidyverse.html)
  - [Chapter 5: Importing data](https://rafalab.github.io/dsbook/importing-data.html)
  
### Git/GitHub and R Markdown:

- In this course we'll be using the version control software [Git](https://en.wikipedia.org/wiki/Git) and its web-based hosting and collaborative platform [GitHub](https://en.wikipedia.org/wiki/GitHub). 
- [The online resource "Happy Git and GitHub for the useR" from Jenny Bryan](https://happygitwithr.com/) is a great reference for these tools as we learn them.
- Another helpful git resource is [Hadley Wickham's webinar "Collaboration and time travel- version control with git, github and RStudio"](https://rstudio.com/resources/webinars/collaboration-and-time-travel-version-control-with-git-github-and-rstudio/)
- We'll learn about using [R markdown](https://rmarkdown.rstudio.com/) to generate readable and reproducible reports with code and text, and you'll be using that a lot in this course - see [Chapter 18 of the 'Happy Git' resource: "Test drive R markdown"](https://happygitwithr.com/rmd-test-drive.html).

## Evaluation

You will have three individual assignments, six seminar submissions (one divided into two parts), and one group project. <font color="red">**Deadlines are all by 11:59 pm (Pacific time) on the due date. Any submission or modification after the due date will not be graded unless you have requested an extension**</font>. If you anticipate having trouble meeting a deadline and need to request an extension/academic concession please reach out via email.

**For more detail on each of these assignments, see the [assignments page](assignments.html)** (the header of each assignment on this page points to the relevant section of the assignments page). Also refer to this [visual overview of the timeline](timeline.html).

**For detailed instructions on how to work and submit assignments using GitHub for this course, please see the [Submission Guide](submission_guide.html).**

### [Intro Assignment (5%)](assignments.html#intro-assignment-5)

- An introductory assignment designed to assess basic knowledge of GitHub, R and Rmarkdown

```{r, echo = FALSE, warning =  FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
assn <- read.csv("course-admin/assn_2022.csv")
assn <- assn %>% 
  select(content,start,Category) %>%
  rename(Assignment = content, `Due Date` = start) %>%
  mutate(`Due Date` = as.Date(`Due Date`, "%Y-%m-%d")) %>%
  arrange(`Due Date`) %>% 
  mutate(`Due Date` = format(`Due Date`, "%a %d %B %Y"))
kable(assn %>% filter(Assignment == "Intro Assignment") %>% select(-Category))
```

### [Seminar completion (10%)](assignments.html#seminar-completion-10)

- You will submit short "deliverables" for seminars 1, 2 (split into two parts - Seminar 2a and 2b together count as one deliverable), 3, 4, 5, and 7
- Each of these six Seminar session deliverables is weighted equally, but the lowest score will be dropped (so that the 5 deliverables with highest score will each count for 2% of the final grade). 
- These deliverables give practical experience applying the knowledge that will be helpful on the homework assignment, final project, and (hopefully) your future research.
- Each deliverable is due on the Friday following the TA-led session for that seminar


```{r, echo = FALSE, warning =  FALSE}
kable(assn %>% filter(Category == "Seminar") %>% select(-Category))
```

### [Paper critique (5%)](assignments.html#paper-critique-5)

- Read, summarize and critique [this paper](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0159197)
- For detailed instructions, refer to [the rubric](critique.html)

```{r, echo = FALSE, warning =  FALSE}
kable(assn %>% filter(Category == "Paper Critique") %>% select(-Category))
```

### [Analysis assignment (30%)](assignments.html#analysis-assignment-30)

- Involves detailed analysis of real data using R
- This assignment will assess your ability to understand and apply the methods learned in class 

```{r, echo = FALSE, warning =  FALSE}
kable(assn %>% filter(Category == "Individual Assignment", Assignment != "Intro Assignment") %>% select(-Category))
```

### [Group project (50%)](assignments.html#final-group-project-50) 

- A semester-long data analysis group project that will allow you to apply the techniques covered in class to a research question of your choosing
- Groups of target size of 4 students will be formed at the beginning of the course
- Important checkpoints during the term (with deliverables): 
    - Initial project proposal (one-paragraph)
    - Finalize the one page project proposal 
    - Progress report
    - 15 minute oral presentation 
    - GitHub repository
    - Individual report deadline
- For more details on the project components and how groups are selected see the [assignments page](assignments.html#final-group-project-50)
- For detailed grading rubrics of the final project components, see the [final project rubric page](group_project_rubrics.html) and the [final project presentation rubric](course-admin/ProjectPresentationRubric.pdf)


```{r, echo = FALSE, warning = FALSE}
kable(assn %>% filter(Category == "Group Project") %>% select(-Category))
``` 

## Academic Concession {#concession}

If you anticipate having trouble meeting a deadline and need an [academic concession](https://students.ubc.ca/enrolment/academic-learning-resources/academic-concessions), please reach out in advance via email to the instructors. [Here](course-admin/concession_template.pdf) is a template you can use for a self-declaration.

## Academic Integrity

Not only is academic integrity is essential to the successful functioning of our course, but adopting best practices will benefit you in your research practice. Make sure you understand UBC’s definitions of academic misconduct and its consequences. Policy guidelines can be found [here](http://www.calendar.ubc.ca/Vancouver/index.cfm?tree=3,54,111,958).

What does academic integrity look like in this course? 

* **Do your own work**. All individual work that you submit should be completed by you and submitted by you. Do not receive or share completed coursework with students who take the course in another term.

* **Acknowledge others’ ideas**. Scholars build on the work of others, and give credit accordingly. This refers to both outside sources (e.g. from the literature), and inside sources (e.g from your peers).

* **Learn to avoid unintentional plagiarism.** Visit the [Learning Commons’ guide to academic integrity](http://learningcommons.ubc.ca/resource-guides/avoiding-plagiarism/) to help you organize your writing as well as understand how to prevent unintentional plagiarism. 

At any time: if you are unsure if a certain type of assistance is authorized, please ask us.  

## Video etiquette for online classes 

During online lecture and seminar sessions, please observe the following guidelines to help foster a productive learning environment:

* Please keep your microphone muted when you aren't speaking to minimize background noise 
* To minimize 'Zoom fatigue', you may keep your camera off when you are not speaking. If your environment is suitable and you feel comfortable doing so, please turn your camera on when speaking. 
* Please make use of the chat feature to ask questions or make comments - the TAs will be monitoring this space.

## Privacy

#### Lecture streaming & recording

Zoom lectures, in person classroom video streams, and their recordings are only provided to students registered in the course. To protect the privacy of fellow students, please do not share these links or files with others.

#### GitHub

This course requires students to work and communicate on github.com. Please be advised that the material you put on GitHub will be stored on servers outside of Canada.

## Learning Environment & Support

We strive to provide a learning environment where all students can succeed. Please join me in contributing to a classroom culture where everyone feels valued. If you encounter an issue that presents a barrier to your learning, please reach out to us. You can also contact the Ombudsperson for help: https://ombudsoffice.ubc.ca. If you have a documented disability that affects your learning, you may contact the UBC Centre for Acessibility: https://students.ubc.ca/about-student-services/centre-for-accessibility, and contact us as soon as possible if you think you may require accommodation options for course work.

Your mental health and wellbeing impacts your academic performance. Sometimes it is possible to manage challenges on your own, while other times you may need support. UBC is committed to providing student mental health and wellbeing resources, strategies, and services that help you achieve your goals. Visit https://students.ubc.ca/health for more information.

## Covid Safety in the Classroom

### Masks

Masks are required for all indoor public spaces on campus, including classrooms, as per the BC Public Health Officer orders and UBC policy. For our in-person meetings in this class, it is important that all of us feel as comfortable as possible engaging in class activities while sharing an indoor space. For the purposes of this order, the term "masks" refers to medical and non-medical masks that cover our noses and mouths. Masks are a primary tool to make it harder for COVID-19 to find a new host. You will need to wear a medical or non-medical mask for the duration of our class meetings, for your own protection, and the safety and comfort of everyone else in the class. Your mask should cover your nose and mouth. Please do not eat in class. If you need to drink water/coffee/tea/etc, please keep your mask on between sips. 

Students who need to request an exemption to the indoor mask mandate must do so based on one of the grounds for exemption detailed in the [PHO Order on Face Coverings (COVID-19)](https://www2.gov.bc.ca/assets/gov/health/about-bc-s-health-care-system/office-of-the-provincial-health-officer/covid-19/covid-19-pho-order-face-coverings.pdf). Such requests must be made through the Center for Accessibility (info.accessibility@ubc.ca). 

Mask wearing protects you as well as others in your environment. Let's do everything we can as a community to stop the spread of this virus.

### Vaccination

If you have not yet had a chance to get vaccinated against COVID-19, vaccines are available to you, free: http://www.vch.ca/covid-19/covid-19-vaccine. The higher the rate of vaccination in our community overall, the lower the chance of spreading this virus. You are an important part of the UBC community. Please arrange to get vaccinated if you have not already done so. Please also arrange to receive your booster shot once you are eligible, to maximize your protection against emerging variants.

### Seating in class

To reduce the risk of COVID-19 transmission, please sit in a consistent area of the classroom each day. This will minimize your close contacts and will still allow for the pedagogical methods planned for this class to help your learning.

### Your personal health

If you're sick, it's important that you stay home – no matter what you think you may be sick with (e.g., cold, flu, other).

* A daily self-health assessment is required before attending campus. Every day, before coming to class, complete the self-assessment for Covid symptoms using this tool: https://bc.thrive.health/covid19/en
* Do not come to class if you have symptoms of COVID-19, have recently tested positive for COVID-19, or are required to self-isolate or quarantine. You can check this website to find out if you should self-isolate or self-monitor: http://www.bccdc.ca/health-info/diseases-conditions/covid-19/self-isolation#Who.
* If you came to class prior to developing symptoms but later test positive for COVID-19, you are not required to inform your classmates. Public Health conducts contact tracing of positive COVID-19 cases to identify ["close contacts"](http://www.bccdc.ca/health-info/diseases-conditions/covid-19/self-isolation/close-contacts), and will identify and inform any individuals at risk. 
* Your precautions will help reduce risk and keep everyone safer. In this class, the marking scheme is intended to provide flexibility so that you can prioritize your health and still be able to succeed.

What to do if you miss class due to self-isolation requirements or any illness:

* View streaming or recorded lectures, if available
* Consult the class resources on the course website
* Use the class Discussion repo on GitHub to carry out discussions
* Come to virtual office hours
* Seek [academic concessions](#concession), as needed

### Instructor health

If we (the instructors or TAs) are sick: If one of us is ill, develops COVID-19 symptoms, or test positive for COVID-19, then we will not come to class on the days we are scheduled to teach. If that happens, here's what you can expect:

* If this happens on a day we are scheduled to be in person, and we are well enough to teach (most likely), we will still conduct online lectures through Zoom until we are well. In this case, we will post an announcement on the class Discussion repo on GitHub with information. You can anticipate that this would very likely be a last minute announcement. Our classroom will still be available for you to sit and attend an online session, although it is recommended that you bring headphones.
* If we are not well enough to teach, then either the other instructor (for the case of lectures), or the other TA (for the case of seminars), will try to be available to fill in. If not, we will try to make new resources to aid in your learning, or make accommodations regarding the assignments.

